\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-20 doc/incertae_sedis.tex
Interchange 8b and 8c.

However, equation (\ref{eq:exsetformonea}) is not the only set of rules 
for generating a split representation for a taxonomy. For example
consider the case of the taxonomy shown in Figure
\ref{fig:Placing-one-incertae}a. Either of the two matrices shown in
\ref{fig:Placing-one-incertae-MRP}a or
\ref{fig:Placing-one-incertae-MRP}b could represent this taxonomy 
along with its \emph{incertae sedis} interpretation.  The matrix in
Figure \ref{fig:Placing-one-incertae-MRP}a, which corresponds to
equation (\ref{eq:exsetformonea}), assumes that \emph{incertae sedis}
taxa may only intrude into taxa that are siblings in the taxonomy.
The placement of $W$ within $V$ leads to taxon $V$ not being named in
the supertree shown in Figure \ref{fig:Placing-one-incertae}c under
the encoding in \ref{fig:Placing-one-incertae-MRP}a because
$V$ is not a sibling of $W$ in Figure \ref{fig:Placing-one-incertae}a.

The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b allows
$V$ to be named in the supertree shown in Figure
\ref{fig:Placing-one-incertae}c because it removes the assumption that
\emph{incertae sedis} taxa can only be placed within siblings.
However, as a side-effect, if taxon $WX$ is found to be
non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}b
allows $W$ to be placed within $V$.  In general, this approach would
allow incertae sedis children of any taxon $\Gamma_1$ to be placed
arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
that \emph{incertae sedis} taxa 
can only be placed within siblings, and therefore does not know what
to do in this situation.  Thus, it is not clear that either Figure
\ref{fig:Placing-one-incertae-MRP}a or Figure
\ref{fig:Placing-one-incertae-MRP}b comprise a unique, best
encoding of the taxonomy.  Thus, to honor the input taxonomic content
accurately, we may need more taxonomic information than simply a
hierarchy and knowledge of which taxa can float more tipward.

Note that equation (\ref{eq:exsetformonea}) allows \emph{incertae sedis}
taxa to intrude only into siblings in the original taxonomy, and not
into taxa that become siblings through solution to the supertree
problem. This can lead to cases where computing a supertree from two
input trees $T_1$ and $T_2$ and an original taxonomy $\taxonomy$ can
lead to a different outcome than if we compute a supertree from $T_1$
and $\taxonomy$, and then compute a supertree from $T_2$ and a revised
taxonomy $\taxonomy^\prime$. Consider a case in which the taxonomy
$\taxonomy$ is given in Figure \ref{fig:Placing-one-incertae}a (with
matrix representation given by \ref{fig:Placing-one-incertae-MRP}a)
and $T_1$ and $T_2$ each provide the single rooted splits $V1\,X \mid \bullet\,U$ and $W\, V1 \mid
\bullet\,V2$ respectively. 
Running the supertree pipeline on $T_1$  and $\taxonomy$ yields the tree in
Figure \ref{fig:Placing-one-incertae}b.  We use this tree as the revised
taxonomy $\taxonomy^\prime$, and its matrix representation appears in
Figure \ref{fig:Placing-one-incertae-MRP}c.  Running the supertree
pipeline on $T_2$  and $\taxonomy^\prime$ then
results in the tree in Figure \ref{fig:Placing-one-incertae}c.

One might therefore hope that running the supertree pipeline on $T_1$,
$T_2$, and $\taxonomy$ simultaneously would yield the same result.
While the topology for the final tree would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
    or not to name $V$ differs.  As expected,  the definition of $WX$
    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
    storing the revised taxonomy with an \emph{incertae sedis}
    annotation and then reprocessing the taxonomy using our rooted
    partial splits approach would also shift the definition of $V$
    from a taxon that must exclude $W$ to one that may include $W$. 
    This results from the fact that equation (\ref{eq:exsetformonea})
    only allows \emph{incertae sedis} taxa to intrude into taxa that
    are their
    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}b
    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
    which solves the seeming inconsistency.  However, this has some
    surprising side-effects as mentioned above, and is therefore not
    the approach that we present in this paper.

@@

\caption{\label{fig:Placing-one-incertae}
  Placement of incertae sedis taxa within non-siblings could occur if
  taxa become siblings by placement.  Given the taxonomy (a),
  placement of $V$ within $WX$ is shown in (b), followed by placement
  of $W$ within $V$ as shown in (c).  Since $W$ is not a sibling of
  $V$ in (a), the method proposed in this paper and illustrated in Figure
  \ref{fig:Placing-one-incertae-MRP}a does not preserve the taxon name
  for $V$.  However, with an alternative method for computing reduced
  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}b,
  the name $V$ would be preserved.
  }
\end{figure}

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill{}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill
\subfloat[from revised taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & 1 \\
$V2$ & 1 & 1 \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
\end{figure}

>>>>

However, equation (\ref{eq:exsetformonea}) is not the only set of rules 
for generating a split representation for a taxonomy. For example
consider the case of the taxonomy shown in Figure
\ref{fig:Placing-one-incertae}a. Either of the two matrices shown in
\ref{fig:Placing-one-incertae-MRP}a or
\ref{fig:Placing-one-incertae-MRP}c could represent this taxonomy 
along with its \emph{incertae sedis} interpretation.  The matrix in
Figure \ref{fig:Placing-one-incertae-MRP}a, which corresponds to
equation (\ref{eq:exsetformonea}), assumes that \emph{incertae sedis}
taxa may only intrude into taxa that are siblings in the taxonomy.
The placement of $W$ within $V$ leads to taxon $V$ not being named in
the supertree shown in Figure \ref{fig:Placing-one-incertae}c under
the encoding in \ref{fig:Placing-one-incertae-MRP}a because
$V$ is not a sibling of $W$ in Figure \ref{fig:Placing-one-incertae}a.

The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}c allows
$V$ to be named in the supertree shown in Figure
\ref{fig:Placing-one-incertae}c because it removes the assumption that
\emph{incertae sedis} taxa can only be placed within siblings.
However, as a side-effect, if taxon $WX$ is found to be
non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}c
allows $W$ to be placed within $V$.  In general, this approach would
allow incertae sedis children of any taxon $\Gamma_1$ to be placed
arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
that \emph{incertae sedis} taxa 
can only be placed within siblings, and therefore does not know what
to do in this situation.  Thus, it is not clear that either Figure
\ref{fig:Placing-one-incertae-MRP}a or Figure
\ref{fig:Placing-one-incertae-MRP}c comprise a unique, best
encoding of the taxonomy.  Thus, to honor the input taxonomic content
accurately, we may need more taxonomic information than simply a
hierarchy and knowledge of which taxa can float more tipward.

Note that equation (\ref{eq:exsetformonea}) allows \emph{incertae sedis}
taxa to intrude only into siblings in the original taxonomy, and not
into taxa that become siblings through solution to the supertree
problem. This can lead to cases where computing a supertree from two
input trees $T_1$ and $T_2$ and an original taxonomy $\taxonomy$ can
lead to a different outcome than if we compute a supertree from $T_1$
and $\taxonomy$, and then compute a supertree from $T_2$ and a revised
taxonomy $\taxonomy^\prime$. Consider a case in which the taxonomy
$\taxonomy$ is given in Figure \ref{fig:Placing-one-incertae}a (with
matrix representation given by \ref{fig:Placing-one-incertae-MRP}a)
and $T_1$ and $T_2$ each provide the single rooted splits $V1\,X \mid \bullet\,U$ and $W\, V1 \mid
\bullet\,V2$ respectively. 
Running the supertree pipeline on $T_1$  and $\taxonomy$ yields the tree in
Figure \ref{fig:Placing-one-incertae}b.  We use this tree as the revised
taxonomy $\taxonomy^\prime$, and its matrix representation appears in
Figure \ref{fig:Placing-one-incertae-MRP}b.  Running the supertree
pipeline on $T_2$  and $\taxonomy^\prime$ then
results in the tree in Figure \ref{fig:Placing-one-incertae}c.

One might therefore hope that running the supertree pipeline on $T_1$,
$T_2$, and $\taxonomy$ simultaneously would yield the same result.
While the topology for the final tree would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
    or not to name $V$ differs.  As expected,  the definition of $WX$
    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
    storing the revised taxonomy with an \emph{incertae sedis}
    annotation and then reprocessing the taxonomy using our rooted
    partial splits approach would also shift the definition of $V$
    from a taxon that must exclude $W$ to one that may include $W$. 
    This results from the fact that equation (\ref{eq:exsetformonea})
    only allows \emph{incertae sedis} taxa to intrude into taxa that
    are their
    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}c
    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
    which solves the seeming inconsistency.  However, this has some
    surprising side-effects as mentioned above, and is therefore not
    the approach that we present in this paper.

@@

\caption{\label{fig:Placing-one-incertae}
  Placement of incertae sedis taxa within non-siblings could occur if
  taxa become siblings by placement.  Given the taxonomy (a),
  placement of $V$ within $WX$ is shown in (b), followed by placement
  of $W$ within $V$ as shown in (c).  Since $W$ is not a sibling of
  $V$ in (a), the method proposed in this paper and illustrated in Figure
  \ref{fig:Placing-one-incertae-MRP}a does not preserve the taxon name
  for $V$.  However, with an alternative method for computing reduced
  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}c,
  the name $V$ would be preserved.
  }
\end{figure}

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[from original taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill{}
\subfloat[from revised taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & 1 \\
$V2$ & 1 & 1 \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill
\subfloat[alternative encoding]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Panels (a) and (c) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$. Panel (b) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
\end{figure}
